---
title: "Assignment 1 for Reproducible research"
author: "Tanja Molle"
date: "17 März 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Loading and processing the data

```{r}
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
```


## What is the mean total number of steps taken per day?
To show the total number of steps per day in a histogram, the sum of all steps per day is calculated for each day.
```{r}
sum_steps <- with(data,tapply(steps,date,sum,na.rm=TRUE))
```

```{r sum_steps, echo=FALSE}
hist(sum_steps, xlab = "Total steps", main = "Total steps per day", col="grey")
```
From the sum of all steps per day, we can easily calculate the mean and nedian steps per day.
```{r}
print(mean_steps <- mean(sum_steps))
print(median_steps <- median(sum_steps))
```

## What is the average daily activity pattern?
For this task we need to calculate the averager steps per interval.
```{r}
avg_steps <- with(data, tapply(steps,interval,mean,na.rm=TRUE))
```

```{r avg_steps, echo=FALSE}
plot(avg_steps, main = "Average steps per interval", xlab = "intervals", ylab = "Average steps",type="l")
```

This result shows the interval with the most steps on average and the average steps taken in this interval:

```{r}
print(avg_steps[which.max(avg_steps)])
```

## Imputing missing values
The amount of missing values is a calculation of all rows from this data set subtracted by all complete cases.
```{r}
print(nrow(data) - sum(complete.cases(data)))
```

Now we insert all the missing values with the mean value of that interval over all days.
```{r}
data_avg <- data
data_avg[is.na(data_avg$steps),]$steps <- mean(data_avg$steps,interval = data_avg$interval, na.rm=TRUE)

sum_steps_avg <- with(data_avg,tapply(steps,date,sum,na.rm=TRUE))
```
And this compares the total amount of steps per interval with imputed values and with removed NAs. Imputed NAs lead to more total steps.
```{r}
hist(sum_steps_avg, xlab = "Total steps", main = "Total steps per day", col="black")
hist(sum_steps, xlab = "Total steps", main = "Total steps per day", col="grey",add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```
And again from the steps with imputed values we can easily calculate the mean and median amount of steps per interval. They are now the same instead of very close.
```{r}
print(mean_steps_avg <- mean(sum_steps_avg))
print(median_steps_avg <- median(sum_steps_avg))
```

## Are there differences in activity patterns between weekdays and weekends?

Before we can ploit, we need to divide the data into weekday data and weekend data.
```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Montag", "Dienstag", "Mittwoch", "Donnerstag", "Freitag"))
        return("weekday")
    else if (day %in% c("Samstag", "Sonntag"))
        return("weekend")
    else
        stop("invalid date")
}
data_avg$date <- as.Date(data_avg$date)
data_avg$day <- sapply(data_avg$date, FUN=weekday.or.weekend)
```


With this plot we show the difference in the step patterns between weekdays and weekends.
```{r}
averages <- aggregate(steps ~ interval + day, data=data_avg, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("Intervals") + ylab("Average number of steps")
```
